---
title: "2023 Cherry Blossom Prediction"
author: "Sarah Boyt"
date: "02/28/2023"
output:
  html_document:
    df_print: kable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, error = FALSE, 
                      message = FALSE,
                      fig.align = 'center',
                      out.width = '80%')
```

## Instructions
This model for predicting cherry blossom dates uses the following packages:tidyverse,rnoaa and lubridate. 
```{r, eval=FALSE}
# install packges
install.packages('tidyverse')
install.packages(rnoaa)
install.packages(lubridate)
```


```{r}
#load packages
library(tidyverse)
library(rnoaa)
library(lubridate)
```


## Loading the data


```{r}
# upload cherry bloom data
cherry <- read.csv("data/washingtondc.csv") %>% 
  bind_rows(read.csv("data/liestal.csv")) %>% 
  bind_rows(read.csv("data/kyoto.csv"))
```

## Visualizing the time series
```{r, fig.width=8, fig.height=3, out.width='100%', fig.cap="Time series of peak bloom of cherry trees since 1880 at three different sites."}
cherry %>% 
  filter(year >= 1880) %>%
  ggplot(aes(x = year, y = bloom_doy)) +
  geom_point() +
  geom_step(linetype = 'dotted', color = 'gray50') +
  scale_x_continuous(breaks = seq(1880, 2020, by = 20)) +
  facet_grid(cols = vars(str_to_title(location))) +
  labs(x = "Year", y = "Peak bloom (days since Jan 1st)")
```

## Loading in temperture data and calculating gdd:
The gdd is the accummulated growing degrees on the historical average bloom_doy. 
 Growing degree days are calculated as follows :(Tmax-Tmin)/2 (the baseline is zero degrees celsius)


```{r}
# Find average DOY for each location 
washington_doy <- cherry %>% filter(location == 'washingtondc') %>% filter(year >= 1980) 
 washington_doy<- mean(washington_doy$bloom_doy)
 
 kyoto_doy <- cherry %>% filter(location == 'kyoto') %>% filter(year >= 1980) 
 kyoto_doy<- mean(kyoto_doy$bloom_doy)
 
 liestal_doy <- cherry %>% filter(location == 'liestal') %>% filter(year >= 1980) 
 liestal_doy<- mean(liestal_doy$bloom_doy)

# Take the average of other doys to get estimate for vancouver 
vancouver_doy <- sum(kyoto_doy,washington_doy,liestal_doy)/3
```


```{r}
get_temperature_daily <- function (stationid) {
  ghcnd_search(stationid = stationid, var = c("tmin","tmax",'prcp'), 
               date_min = "1980-01-01", date_max = '2023-02-26') %>%
     reduce(left_join) %>%
    transmute(year = parse_number(format(date, "%Y")),
               date, tmax = tmax / 10, tmin = tmin / 10,
             gd = (tmax + tmin) / 2 - 0, prcp= prcp/10)
}
# find gdd 
historic_temperatures_washington <-
  tibble(location = "washingtondc", get_temperature_daily("USC00186350")) %>%
  filter(!is.na(gd) & gd >=0, !is.na(prcp)) %>% group_by(location,year) %>% 
  mutate(doy = yday(date)) %>% filter(doy <= round(washington_doy)) %>%
  summarize(gdd=sum(gd),prcp=sum(prcp))

historic_temperatures_liestal <-   
   (tibble(location = "liestal", get_temperature_daily("GME00127786"))) %>%
  filter(!is.na(gd) & gd >=0 & !is.na(prcp)) %>% group_by(location,year) %>% 
  mutate(doy = yday(date)) %>% filter(doy <= round(liestal_doy)) %>%
  summarize(gdd=sum(gd),prcp=sum(prcp))

 historic_temperatures_kyoto<-
(tibble(location = "kyoto", get_temperature_daily("JA000047759"))) %>%
   filter(!is.na(gd) & gd >=0 & !is.na(prcp)) %>% group_by(location,year) %>% 
   mutate(doy = yday(date)) %>% filter(doy <= round(kyoto_doy)) %>%
   summarize(gdd=sum(gd),prcp=sum(prcp))
   
   historic_temperatures_vancouver <- 
     (tibble(location = "vancouver", get_temperature_daily("CA001108395"))) %>%
     filter(!is.na(gd) & gd >=0 & !is.na(prcp)) %>% group_by(location,year) %>% 
     mutate(doy = yday(date)) %>% filter(doy <= round(vancouver_doy)) %>%
     summarize(gdd=sum(gd),prcp=sum(prcp))

historic_temperatures <- bind_rows(historic_temperatures_washington,historic_temperatures_kyoto,
                          historic_temperatures_liestal,historic_temperatures_vancouver)

historic_temperatures %>%
  ggplot() + 
  aes(year, gdd) + 
  geom_line(aes(color=location)) +
  xlim(1980, 2022) +
  labs(x = "Year", y = "GDD °C") 

historic_temperatures %>%
  ggplot() + 
  aes(year, prcp) + 
  geom_line(aes(color=location)) +
  xlim(1980, 2022) +
  labs(x = "Year", y = "Precipation mm") 
```



```{r}

## extrapolate gdd and prcp for future years
#### gdd 
ls_fit_gdd <- lm(gdd ~ year*location, 
                         data = historic_temperatures)

summary(ls_fit_gdd)

gdd_predictions <-
  expand_grid(location = c("washingtondc", "liestal", "kyoto", "vancouver" ),
              year = 2023:2032) %>%
  bind_cols(gdd = 
              predict(ls_fit_gdd, newdata = .)) 

#### prcp 
ls_fit_prcp <- lm(prcp ~ year*location,
                  data=historic_temperatures)
summary(ls_fit_prcp)

prcp_predictions <-
  expand_grid(location = c("washingtondc", "liestal", "kyoto", "vancouver" ),
              year = 2023:2032) %>%
  bind_cols(prcp = 
              predict(ls_fit_prcp, newdata = .)) 
cherry_gdd <- historic_temperatures %>% 
  left_join(cherry,
            by=c('year','location')) 

### model 
ls_fit_blossom<- lm(bloom_doy ~ gdd*location+prcp, 
     data = cherry_gdd)
summary(ls_fit_blossom)

### extrapolated data used to predict bloom_doy
gdd_prcp_pred <- left_join(gdd_predictions,prcp_predictions,
                           by=c('year','location'))


### predictions for kyoto, washington dc, and liestal 
prediction <-expand_grid(City = c("washingtondc", "liestal", "kyoto"),
                        Year = 2023:2032) %>%
  bind_cols(predicted_bloom_doy= 
              predict(ls_fit_blossom, newdata = gdd_prcp_pred[1:30,])) 


```

## Extrapolating to Vancouver, BC

```{r}
#vancouver prediction 
ls_fit_blossom_van<- lm(bloom_doy ~ gdd+prcp+ long + lat, 
                    data = cherry_gdd)

summary(ls_fit_blossom_van)

gdd_prcp_pred$lat <- ifelse(gdd_prcp_pred$location=='vancouver',49.22,NA)
gdd_prcp_pred$long <- ifelse(gdd_prcp_pred$location=='vancouver',-123.16,NA)


prediction_van <-expand_grid(City = c("vancouver"),
                         Year = 2023:2032) %>%
  bind_cols(predicted_bloom_doy= 
              predict(ls_fit_blossom_van, newdata = gdd_prcp_pred[31:40,])) 

```

Not surprisingly, the predicted peak bloom dates for Vancouver are thus very similar to the other 3 sites:
```{r}

```

We can check the predictions against the data from previous competition years.
```{r}

```

If satisfied with the predictions, we can append them to the predictions for the other sites:
```{r}
predictions <-prediction %>% arrange(City) %>%
  bind_rows( prediction_van)
```

## Preparing the submission file

Once we have the predictions for all four sites, we have to save them in the correct format for the competition.

```{r}

submission_predictions <- predictions %>% 
  mutate(predicted_bloom_doy = round(predicted_bloom_doy)) %>% 
  pivot_wider(names_from = 'City', values_from = 'predicted_bloom_doy')
submission_predictions
```

For submission, these predictions must be saved as a CSV file.
**Important:** the CSV file must not have row names, which R adds by default. Specify `row.names=FALSE` to suppress them:

```{r, eval=FALSE}
write.csv(submission_predictions, file = "cherry-predictions.csv",
          row.names = FALSE)
```

